\documentclass[fontsize=10pt,twoside,BCOR=12mm,DIV=calc]{scrbook}
\input{preamble}
%%%%%%%%%%%%
% document %
%%%%%%%%%%%%

\begin{document}

%%%%%%%%%
% title %
%%%%%%%%%


\title{\textbf{\
Mathematics
}\\An undergraduate notebook}
\author    {Nathan Typanski}
\date      {\today}
\maketitle

\newpage

\cleardoublepage

\newpage

\edef\marginnotetextwidth{\the\textwidth}

\begin{center}
This work is Copyright \copyright\ 2013 by Nathan Typanski.
\end{center}

This work is made available under the terms of the \emph{Creative Commons
    Attribution-NonCommercial-ShareAlike 3.0 Unported License}.
\textbf{You are free:}
\begin{itemize}
  \item[] \textbf{to Share}\dash{}to copy, distribute and transmit the work
  \item[] \textbf{to Remix}\dash{}to adapt the work
\end{itemize}

\textbf{Under the following conditions:}
\begin{itemize}
  \item[]\ccby\textbf{Attribution}\dash{}You must attribute the work in the manner
      specified by the author or licensor (but not in any way that suggests that
      they endorse you or your use of the work).
  \item[] \ccnc\textbf{Noncommercial}\dash{}You may not use this work for
      commercial purposes.
  \item[] \ccsa\textbf{Share Alike}\dash{}If you alter, transform, or build upon
      this work, you may distribute the resulting work only under the same or
      similar license to this one.
\end{itemize}

\textbf{With the understanding that:}
\begin{itemize}
  \item[]\textbf{Waiver}\dash{}Any of the above conditions can be \emph{waived} if
      you get permission from the copyright holder.
  \item[]\textbf{Public Domain}\dash{}Where the work or any of its elements is in
      the \emph{public domain} under applicable law, that status is in no way
      affected by the license.
  \item[]\textbf{Other Rights}\dash{}In no way are any of the following rights
      affected by the license:
  \begin{itemize}
    \item Your fair dealing or \emph{fair use} rights, or other applicable
        copyright exceptions and limitations;
    \item The author's \emph{moral} rights;
    \item Rights other persons may have either in the work itself or in how the
        work is used, such as \emph{publicity} or privacy rights.
  \end{itemize}
  \item[]\textbf{Notice}---For any reuse or distribution, you must make clear
      to others the license terms of this work.
\end{itemize}
\begin{center}
  \url{http://creativecommons.org/licenses/by-nc-sa/3.0/}

  \cc{}
\end{center}
\include{preface}

\setcounter{tocdepth}{3}
\tableofcontents

\listoftables

\listoffigures

\include{discrete/main}
\include{continuous/main}
%\include{physics/main}
\part{Appendix}
\include{appendix/main}
%\input{discrete/algorithms}

\printindex

% Should we be printing unused bibliography references in these drafts?
% probably not.
%\nocite{*}
\bibliographystyle{plainnat}
\bibliography{Bibliography}
%% ----------------------------------------------------------------
\end{document}
