\chapter*{Preface}\epigraph{
  Hold to the now, the here, through which all future plunges to the
  past.
}
{Ulysses, James Joyce, Episode 9}
\addcontentsline{toc}{chapter}{Preface}

I started this notebook my Math 240 (Calculus II) class at Christopher Newport
University on \date{February 13, 2012}. I had picked up the basics of \LaTeX\ in my free hour
before class because I wanted to learn how to type mathematical documents. Why?
Because \LaTeX\ is cool.

I started to care about math because it serves as the logical foundation for
physics. I realized quickly in my studies that I did not know enough math, and I
did not know it rigorously enough to truly understand physics. I cared about
physics because it is a prerequisite for understanding the basics of computer
engineering, my major.
% thanks Kyle Martin for correcting a typo here.

% EDIT: REMOVED 11/6/2012
%%%%%%%%%%
% Discrete mathematics, a subject which has only recently grown to popularity in
% concurrence with computer science, is included first because it describes much
% of the logical foundation for mathematics in ways I had never encountered
% before. In many ways, it involves thinking about the basic thought processes
% that we take for granted in continuous mathematics. When we make claims in
% mathematics such as
% \begin{quote}
%   ``\(x=6\)''
% \end{quote}
% \begin{quote}
%   ``A limit of sums is a sum of limits.''
% \end{quote}
% there is an underlying logical structure that governs the meanings of such
% statements and how we conceptualize and work with them. For this reason, I have
% found discrete mathematics extremely enlightening in my own study of calculus
% and beyond.

This is now the longest document I've ever written. It has grown to represent
a sizable portion of my college education at this time. It's also the first time I've
developed a sustainable organizational system for my notes. Everything before
this, and everything besides this, lies in stacks of scattered legal pads in at
least four different locations.

My goal is to finally organize my thoughts and conceptualize this material in a
way I have never even attempted before.

Whatever it takes.

\hfill{Nathan Typanski}

\hfill \date{April 9, 2012}
\newpage
\section*{A note to the reader}
This text is a work in progress.

Everything in this document is subject to change.

No claim is made as to the accuracy of any of the information contained herein. There may be mistakes, inaccuracies, or outright lies included among otherwise relevant and complete content. Always check with a reputable source (e.g.\ a math book).

As of \date{July 14, 2013}, I am officially declaring this project to emphasize quantity and inclusion of my old, handwritten notes over quality.
This is part of an effort to insert enough content that this text actually becomes a valuable resource.
Right now, while organization, perfection, and copywrighting would be an admirable goal, there is simply not enough material to ``see the bigger picture,'' as it were.
Therefore, my direction will shift to attack that goal.
I was most productive on this when I \emph{was} mindlessly adding content, in-class notes, what have you.
Things may always be removed later, but it takes considerable effort to write them in the first place.
\newpage
\section*{A note on references}
Whenever a number is appended to a sentence in brackets ([11], for example), it means the preceeding section has a citation in the bibliography and can be examined in the original source.
This is very commonly used in simple tables that are reproduced here, or in sections of text where the material is not different enough from its source text.

The references for this text are not fully established in the official text.
There are still a number of places where citations may be missing, or provided only by name and without a complete entry in the bibliography.
However, I am regularly going back through this text and completely rewriting sections that have been copied verbatim (for example, a couple of proofs in the appendix are not my work, though I am prudent to say so outright).
This is largely because this material is presented in draft form, and many of the citations are provided in the \LaTeX\ source but not yet finalized to the reader.
Over time, all instances of this will be adequately removed.
