\chapter{Ordinary Differential Equations}
\begin{defn}
  A \textbf{differential equation} is an equation involving derivatives.
\end{defn}
\begin{defn}
  A \textbf{direction field} tells us the slope of a function at any given place.
\end{defn}
\begin{ex}
    In physics, we often define acceleration to be a vector relative to another 
    vector, velocity.
    Here, we will just consider them as scalars for the sake of argument.
    Acceleration is a change in velocity, so
    \begin{equation}
        a = \leib{v}{t},
        \label{eq:acceleration}
    \end{equation}
    where $v$ represents velocity and $t$ represents time.
    Now we integrate both sides of \eref{eq:acceleration} with respect to $t$,
    \begin{equation}
        \int a \ud t = \int \leib{v}{t} \ud t.
        \label{eq:intaccel}
    \end{equation}
    Now, assuming\footnote{This must be explained later, but as a warning: no,
    the $\ud t$ in the derivative operation and the $\ud t$ in the integration 
    operation do not simply cancel.}
    \begin{equation}
        \int \leib{v}{t} \ud t = \int \ud v,
        \label{eq:handwave}
    \end{equation}
    then using \eref{eq:handwave}, we see that
    \begin{equation}
        \int a \ud t = \int \ud v.
        \label{eq:naughtint}
    \end{equation}
    From here we simplify, finding that
    \begin{equation}
        at + c_1 = v + c_2.
        \label{eq:almostvelocity}
    \end{equation}
    The constants in \eref{eq:almostvelocity} are simply constants and may be combined into another constant, $C$.
    Also, the equation may be rearranged to put it in more familiar form, yielding
    \begin{equation}
        v = at + C,
        \label{eq:velocity}
    \end{equation}
    which we recognize as the classical mechanics equation for velocity.
    Integrating once more, and replacing $v$ with the definition of velocity as change in position, we find
    \begin{align}
        \int \leib{x}{t} \ud t &= \int (at + C) \ud t, \nonumber \\
        \int \leib{x}{t} \ud t &= \int at \ud t + \int C \ud t, \nonumber \\
        x + c_3  &= a \frac{t^2}{2} + Ct + c_4. \nonumber \\
        \intertext{Now we may simply combine the constants once more, defining $C_1$ to constitute the difference of $c_4$, and $c_3$,}
        x &= a \frac{t^2}{2} + Ct + C_1. \label{eq:phys_const}
    \end{align}
    \eref{eq:phys_const} may be rewritten in its more common form:
    \begin{equation}
        x(t) = \frac{1}{2} a t^2 + v_0 t + x_0.
        \label{eq:position}
    \end{equation}
\end{ex}

% \section{Integrating Factors}
% We will observe differential equations of the form
% \begin{equation}
%   \leib{y}{t} = g(t) y + r(t)
% \end{equation}
% \begin{enumerate}
%   \item $\leib{y}{t} = t^2 y + \cos{t}$
%   \item $t y +3=\leib{y}{t}-2t$
% \end{enumerate}
