\chapter{Rules of Inference}
  \epigraph{I don't want to believe. I want to know.}
{Carl Sagan}
\label{ch:rules-of-inference}
Proofs are used to establish the truth of mathematical statements. In order to
make a proof, we must use the \textbf{rules of inference}\index{rules of
inference} to establish the truth of more complicated logical arguments. An
\textbf{argument} is a sequence of propositions that ends with a conclusion. A
\textbf{valid} argument is one in which the last proposition follows from those
propositions before it.

When we are writing mathematical proofs, it's not common to actually cite the rules of inference in our text.
However, they should form the logical connectors between the claims we make in our proofs and should be present in the implicit form.
Becoming familiar with these rules, and how to use them, will allow us to both write more coherent proofs and to avoid logic errors in our writing.

\section{Rules of Inference for Propositions}

We will present the rules of inference using a variant of \emph{Fitch diagrams}.
Each step in a Fitch diagram includes a number for the step, a proposition or conclusion,
and a justification for the step.

\subsection{\emph{Modus ponens}}\label{modus_ponens}\index{\emph{modus ponens}}
\begin{equation*}
  \begin{fitch}
    \fb p       & assumption \\
    \fa p \to q & assumption \\
    \fa  q & $\big(p \wedge (p \to q)\big) \to q$
  \end{fitch}
\end{equation*}

Another way you might see this written is

\begin{array}{rl}
1. & p \rightarrow q \\
2. & p \\
\hline
\therefore & q
\end{array}

where the three dots ($\therefore$) is read as ``therefore''.
We will usually try to stick with the fitch diagrams.

\emph{Modus ponens} is Latin for ``mode that affirms,'' and comes from the
tautology $\big(p \wedge (p \to q)\big) \to q$. It is the simplest valid
\textbf{argument}\index{argument}, a sequence of statements that ends with a conclusion.

\subsection{\emph{Modus tollens}}\index{\emph{modus tollens}}
\begin{equation*}
  \begin{fitch}
    \fb \neg q  & assumption \\
    \fa p \to q & assumption \\
    \fa \neg p & $\big(\neg q \wedge (p \to q)\big) \to \neg p$
  \end{fitch}
\end{equation*}

\subsection{Hypothetical syllogism}\index{hypothetical syllogism}
\begin{equation*}
  \begin{fitch}
    \fb p \to q& assumption \\
    \fa q \to r& assumption \\
    \fa p \to r& $\big((p \to q) \wedge (q \to r)\big) \to (p \to r)$
  \end{fitch}
\end{equation*}
A hypothetical syllogism is sometimes thought of as \emph{double modus ponens}.

\subsection{Disjunctive syllogism}\index{disjunctive syllogism}
\begin{equation*}
  \begin{fitch}
    \fb p \vee q & assumption \\
    \fa \neg p   & assumption \\
    \fa q & $(p \lor q ) \land \neg p \to q$
  \end{fitch}
\end{equation*}

\subsection{Addition}\index{addition}
\begin{equation*}
  \begin{fitch}
    \fb p & assumption \\
    \fa p \lor q & $ p \to (p \lor q)$
  \end{fitch}
\end{equation*}

\subsection{Simplification}\index{simplification}
\begin{equation*}
  \begin{fitch}
    \fb p \land q & assumption \\
    \fa p & $(p \land q) \to p$
  \end{fitch}
\end{equation*}

\subsection{Conjunction}\index{conjunction}
\begin{equation*}
  \begin{fitch}
    \fb p & assumption \\
    \fa q & assumption \\
    \fa p \land q & $\big( (p) \land (q)\big) \to (p \land q)$
  \end{fitch}
\end{equation*}

\subsection{Resolution}\index{resolution}
\begin{equation*}
  \begin{fitch}
    \fb p \lor q      & assumption \\
    \fa \neg p \lor r & assumption \\
    \fa q \lor r & $ \big( (p \lor q) \land (\neg p \lor r ) \big) \to (q \lor r)$
  \end{fitch}
\end{equation*}

\section{Rules of Inference for Quantified Statements}

\subsection{Universal Generalization}

\textbf{Universal generalization} states that given $P(c)$ for all elements $c$
in the domain, $\forall x P(x)$ is true.
\begin{equation}
  \begin{fitch}
    \fb P(c) \text{ for some arbitrary $c$} & assumption \\
    \fa \forall x P(x) & universal generalization
  \end{fitch}
  \label{eq:univ_gen}
\end{equation}

\subsection{Universal Instantiation}\label{univ_inst}

\textbf{Universal instantiation}  states that given $\forall x P(x)$, $P(c)$ is
true for a particular element $c$ in the domain.
\begin{equation}
  \begin{fitch}
    \fb \forall x \big(P(x) \to Q(x)\big) & proposition \\
    \fa P(a) & universal instantiation
  \end{fitch}
  \label{eq:univ_inst}
\end{equation}

\subsection{Existential Generalization}

\textbf{Existential generalization}\index{existential generalization} concludes
that, given a particular element $c$ for which $P(c)$ is known to be true, $\exists x P(x)$.

\subsection{Existential Instantiation}

\textbf{Existential instantiation}\index{existential instantiation} states that if
$\exists x P(x)$ is true, $P(c)$ for some element $c$.

\subsection{Universal \emph{Modus Ponens}}

\textbf{Universal \emph{modus ponens}} combines universal instantiation
(Section \ref{univ_inst}) and \emph{modus ponens} (Section \ref{modus_ponens}) to
tell us that if $\forall x (P(x) \to Q(x) )$ is true, and if $P(a)$ is true for a
particular element $a$ in the domain of the universal quantifier, then $Q(a)$ must
also be true.
\begin{equation}
  \begin{fitch}
    \fb \forall x (P(x) \to Q(x)) \\
    \fa P(a), \text{ where $a$ is a particular element in the domain} \\
    \fa Q(a)
  \end{fitch}
  \label{eq:univ_mod_pon}
\end{equation}

\subsection{Universal \emph{Modus Tollens}}

\textbf{Universal \emph{modus tollens}} states that
\begin{equation}
  \begin{fitch}
    \fb \forall x (P(x) \to Q(x)) \\
    \fa \neg Q(a), \text{ where $a$ is a particular element in the domain} \\
    \fa \neg P(a)
  \end{fitch}
  \label{eq:univ_mod_tol}
\end{equation}

