\part{Discrete Mathematics} % All of this part comes from Rosen's Discrete Mathematics and its Applications, Seventh Edition.
\yinipar{\color{Maroon}T}his is the part of this text which won't focus on numbers.
Were it not for specific examples, no numbers would exist in this part altogether.
This part sets the stage for the logical theory behind mathematics. We will start with \emph{propositional logic}, a very simple form of logic that establishes the groundwork for logic statements.
From there, we will describe \emph{predicate logic}, essentially a more powerful variant of propositional logic.
We will then analyze the idea of mathematical \emph{proofs}, eventually working our way toward \emph{set theory}, the groundwork of what most people know as mathematics.

Most generalized ``calculus'' books don't teach mathematics in this order.
\emph{Thomas' Calculus,} which I reference often in this text, dedicates only one page in its appendix to sets.
I do not believe logic systems are mentioned at all.
Personally, I believe set theory is quite relevant as it can be used to describe almost everything in mathematics.
It is often ignored in introductory calculus classes because it is considered so intuitive as to not be important to students.

Set theory will offer us a look at mathematics as a \emph{system}.
It will allow the theory behind math to shine through to the surface, offering a vivid understanding of what goes on beneath the surface of equations, calculations, and variables.
Because of this, we will use the language of set theory in many of our definitions throughout this text.

\setcounter{section}{0}
\input{discrete/propositional}
\input{discrete/predicates}
\input{discrete/inference}
\input{discrete/proofs}
\input{discrete/sets}
\input{discrete/recursion}
\input{discrete/counting}
\input{discrete/numbers}
