\chapter{Important Concepts}

\section{Quadratic Formula}
\index{quadratic formula}
Quadratic formula\index{quadratic formula}
\begin{equation}
  x=\frac{-b\pm\sqrt{b^2-4ac}}{2a}
  \label{app:eq:quadratic}
\end{equation}

\section{Point-slope Formula}
\label{sec:pointslope}
\index{point-slope formula}
The \textbf{point-slope formula} allows us, given a point $(x_1, y_1)$ and a slope $m$, to solve for $y$ as a function of $x$.
\begin{equation}
  \label{eq:pointslope}
  y-y_1=m(x-x_1)
\end{equation}

\section{Conjugate}\label{app:def:conjugate}
\index{conjugate}
In algebra, the \textbf{conjugate}\index{conjugate} of a \emph{binomial} is another binomial formed by taking the opposite of the second term of the first binomial. For the initial binomial
\[ a + b\]
its conjugate would be
\[a - b.\]

Meanwhile, for the expression \[a^2+b^2\] we can factor this to produce \[(a-b)(a+b)\] where one expression is the conjugate of the other.

\chapter{Proofs}

\section{Power Rule for Derivatives}
The \emph{power rule for derivatives}\index{power rule} states that
\begin{equation}
  \label{eq:pwrrlprf}
  \ddx x^n=nx^{n-1}
\end{equation}
\begin{proof}
  To prove this, we use the limit definition of a derivative:
\begin{equation}
  \label{eq:lmtdefprf}
  \frac{\ud f(x)}{\ud x}=\lim_{h \to 0} \frac{f(x+h)-f(x)}{h}
\end{equation}
And assume that $f(x)$ is of the form $f(x)=x^n$.

Look at the term $(x+h)^n$. Take $n=1$. In this case:
\begin{align*}
  (x+h)^1 &= x+h \\
  \intertext{Try it for $n=2$.}
  (x+h)^2 &= (x+h)(x+h) \\
  &= x^2+2hx+h^2 \\
  \intertext{Now for $n=3$.}
  (x+h)^3 &= (x+h)(x+h)^2 \\
  &= (x+h)(x^2+2hx+h^2) \\
  &= x^3 + 2hx^2 + h^2x + hx^2 + 2h^2x + h^3 \\
  &= x^3 +3hx^2 + 3h^2x +h^3
  \intertext{We are beginning to see a pattern in each of these sums: the first term is $x^n$, and each term after that has a common factor of $h$.
    Furthermore, it looks like there is always a term in the sum that has only one $h$ within.
  Let's find this for $n=4$ to be sure:}
  (x+h)^4 &= (x+h)(x+h)^3 \\
  &= (x+h)(x^3+3hx^2+3h^2x+h^3) \\
  &= x^4+3hx^3+3h^2x^2+h^3x+x^3h+3hx^2x^2+3h^3x+h^4 \\
  \intertext{Simplify.}
  (x+h)^4&= x^4 +4hx^3+6h^2x^2+4h^3x+h^3x+x^3h+h^4 \\
\end{align*}
It looks like the pattern follows for $n=4$.
We can claim that this holds for any value $n$, supposing that $f(x)$ is of the form $x^n$.

What does this mean for our limit equation \eqref{eq:lmtdefprf}?
\begin{align*}
  \ddx f(x)&=\lim_{h \to 0} \frac{f(x+h)-f(x)}{h} \\
  \intertext{Well, it means that the first term of our expanded polynomial should always cancel with the $\cdots -f(x)$ term if $n\neq 1$.
    It also means that we should always be able to divide our numerator by $h$ to calculate the limit, assuming, again, that $n\neq 1$.
    Our limit would then look like this:}
    \ddx f(x) &= \lim_{h \to 0} \frac{(x^n+n\cdot h x^{n-1}+\cdots +h^n)-x^n}{n} &n\neq 1\\
    \intertext{$x^n$ terms cancel.}
    &=\lim_{h \to 0} \frac{n\cdot h x^{n-1}\cdots+h^n}{h}&n\neq 1\\
\\
    \intertext{Divide by $h$}
    &=\lim_{h \to 0} nx^{n-1}\cdots+h^{n-1} &n\neq 1, \quad h\neq 0\\
    \intertext{All terms following the first go to $0$ as $h \to 0$, and we are left with}
    \ddx f(x) &=  nx^{n-1}
  \end{align*}
  Which is the same as \eqref{eq:pwrrlprf}.
\end{proof}

\section{Sandwich Theorem for Sequences}

I didn't write any of this. Check the citations for my sources.

\label{proof:sandwichsequence}
\begin{theorem}[The Sandwich Theorem for Sequences]\index{The Sandwich
  Theorem for Sequences}
  \label{app:th:sandwichsequence}
  Let $\{a_n\}$, $\{b_n\}$, and $\{c_n\}$ be sequences of real numbers. If
  $a_n \leq b_n \leq c_n $ holds for all $n$ beyond some index $N$, and if
  $\lim_{n\to\infty} a_n = \lim_{n\to\infty} c_n = L$, then
  $\lim_{n\to\infty} b_n = L$ also.
\end{theorem}
\begin{theorem}
  \label{app:th:xninfty}
  If \(|x| < 1, \lim_{n \to \infty} x^n = 0\).
  \begin{proof}
    We need to show that to each $\varepsilon > 0$ there corresponds an
    integer $N$ so large that $\forall n > N \big( |x^n| < \varepsilon\big)$
    Since $\varepsilon^{1/n}\to 1$, while $|x|<1$, $\exists N \big(
    \varepsilon^{1/n} > |x|$. In other words,
    \begin{equation}\label{app:eq:limxto1}
      \Big|x^N\Big| = \Big|x^N\Big| < \varepsilon.
    \end{equation}
    This is the integer we seek because, if $|x|< 1$, then
    \begin{equation}\label{app:eq:limxto2}
      \forall n > N \bigg( \Big|x^n\Big| < \Big|x^N\Big|\bigg)
    \end{equation}
    Combining \eqref{app:eq:limxto1} and \eqref{app:eq:limxto2} produces
    $|x^n|<\varepsilon$ for all $n > N$, concluding the
    proof.\cite[p.~AP-21]{thomas}
  \end{proof}
\end{theorem}
\label{proof:limxnnf}
\begin{theorem}
  % Theorem and proof are a direct rip from Thomas' Calculus.
  For any number $x$, $\lim_{n\to\infty} \frac{x^n}{n!} = 0$.
    \cite[p.~AP-22]{thomas}
  \begin{proof}
  Since
  \[ - \frac{|x|^n}{n!} \leq \frac{x^n}{n!} \leq \frac{|x|^n}{n!},\]
  all we need to show is that $|x|^n/n! \to 0$. We can then apply theorem
  \ref{app:th:sandwichsequence} to conclude that $x^n/n! \to 0$.

  The first step in showing that $|x|^n/n! \to 0$ is to choose an integer $M
  > |x|$, so that $(|x|/M) < 1$. By Theorem \ref{app:th:xninfty}, we then have
  that $\left( |x|/M \right)^n \to 0$. We then restrict our attention to
  values of $n > M$. For these values of $n$, we can write
  \begin{align*}
    \frac{|x|^n}{n!}&=
    \frac{|x|^n}{1\cdot2\cdot\cdots\cdot M
      \cdot\underbrace{(M+1)\cdot(M+2)\cdot\cdots n}_{(n-M) \text{
      factors}}}
      \\
      &\leq \frac{|x|^n}{M!M^{n-M}}=\frac{|x|^n \cdot M^M}{M^n \cdot M!} =
      \frac{M^M}{M!} \left( \frac{|x|}{M} \right)^n .
  \end{align*}
  Thus,
  \[ 0 \leq \frac{|x|^n}{n!}\leq \frac{M^M}{M!}\left( \frac{|x|}{M}
    \right)^n .\]
    Now, the constant $M^M / M!$ does not change as $n$ increases. Thus
    Theorem \ref{app:th:sandwichsequence} tells us that $|x|^n/n! \to 0$ because
    $(|x|/M)^n \to 0$.
    \cite[p.~AP-22]{thomas}
  \end{proof}
\end{theorem}

\section{L'Hospital's Rule}
\label{proof:lhospital}
\begin{theorem}[L'Hospital's Rule]
  Suppose that $f(a)=g(a)=0$, that $f$ and $g$ are differentiable on an open interval $I$ containing $a$, and that $g'(x) \neq 0$ on $I$ if $x \neq a$. Then
  \[ \lim_{x \to a} \frac {f(x)}{g(x)} \=H \lim_{x \to a} \frac{f'(x)}{g'(x)} \]
  assuming that the limit on the right side of this equation exists.
\end{theorem}
\begin{proof}
  We first establish the limit equation for the case $x \to a^+$. The method needs almost no change to apply to $x \to a^{-}$, and the combination of these two cases establishes the result.

  Suppose that $x$ lies to the right of $a$. Then $g'(x) \neq 0$, and we can apply Cauchy's Mean Value Theorem to the closed interval from $a$ to $x$. This step produces a number $c$ between $a$ and $x$ such that
  $$ \frac{f'(c)}{g'(c)}=\frac{f(x)-f(a)}{g(x)-g(a)} $$
  But $f(a)=g(a)$, so
  $$ \frac{f'(c)}{g'(c)}=\frac{f(x)}{g(x)} $$
  As $x$ approaches $a$, $c$ approaches $a$ because it always lies between $a$ and $x$. Therefore,
  $$ \lim_{x \to a^+} \frac{f(x)}{g(x)}=lim_{c \to a} \frac{f'(c)}{g'(c)} = lim_{x \to a^+} \frac{f'(x)}{g'(x)} $$
  which establishes l'Hospital's Rule for the case where $x$ approaches $a$ from
  above. The case where $x$ approaches $a$ from below is proved by applying
  Cauchy's Mean Value Theorem (found in Section \ref{sec:lhospital}) to the closed interval $[x,a], x <
  a$.\cite{thomas}
\end{proof}
\input{appendix/logic.tex}

\chapter{The Fibonacci Sequence and the Golden Ratio}
\input{fibonacci/fib.tex}
